Simplifying Lotka--Volterra interaction matrices by species grouping: optimal groupings are unpredictable

### 2022-05-17 article.rnw
Remove a sentence (about reading Terry for methods) that DBS commented out.
Remove a clause from the methods that DBS commented out.
Remove a break between two paragraphs.
The second "paragraph" was a single sentence, which is poor practice.
Add a sentence explaining how many parameters the ungrouped model has.
Rearrange the methods section on species groupings.

%We shall not repeat the entire experimental methodology here---refer to \textcite{terry2021natural} for that---but provide a summary of the pertinent points.
% DBS: I don't think we need the above because we say the same thing in line 172 below.

@@

The experiment used six species of \emph{Drosophila} from an Australian rainforest community: \emph{D. birchii}, \emph{D. pallidifrons}, \emph{D. pandora}, \emph{D. pseudoananassae}, \emph{D. simulans}, and \emph{D. sulfurigaster}.
The experiment took place in vials with a limited amount of medium for egg-laying and larval development; competition was for space on this medium.
A fixed number of adult pairs of one or two species were added to each vial and allowed to mate and lay eggs.
The number of adult pairs of each species emerging from these eggs was counted.
%---this was the size of the next generation.
% DBS: Do we need the bit I've commented out above? Plus, it isn't the size of the next generation it's the number of new adults...

Experiments occurred for every single-species and two-species combination of the six species, at a variety of densities.
The single-species densities were 3, 9, or 12 pairs per vial, and each of these was replicated six times per species.
The two-species densities were (3,3), (3,6), (3,9), (3,12), (6,3), (9,3), or (12,3) pairs per vial (where the paired values give the densities of each of the two species), and each of these was replicated three times per pair of species.
Vials with a single founding pair were excluded from this study and the original authors' due to an Allee effect that the Lotka--Volterra model cannot accommodate.
 %
Further details of the experimental methodology can be found in \textcite{terry2021natural}.

@@

The data provided the change in population size from one generation to the next, so we used a discrete-time version of the Lotka--Volterra competition model given by:
\begin{equation}
	N_{i,t+1} - N_{i,t} = N_{i,t} \, r_i \left(1 - \sum_j \alpha_{ij} N_{j,t}\right) \,,
	\label{LVmodel}
\end{equation}
where $N_{i,t}$ is the population of species $i$ at time $t$ (the size of the parent generation), $N_{i,t+1}$ is the population at time $t+1$ (the size of the offspring generation), $r_i$ is the intrinsic growth rate of species $i$, and $\alpha_{ij}$ is the per capita effect of species $j$ on species $i$.
The competition coefficients were not constrained to be positive, allowing for the possibility of mutualistic effects between some species.

% DBS: Let's add something that very briefly outlines how many parameters this model has. Then, below, we can provide formulae for the number of parameters given the number of row and column groups. Even though people can work this out themselves, it is probably useful to make it explicit so that we don't have to depend on them actually mustering up the energy to do so.

@@

% DBS: I still feel like the bits and pieces of the section below are not in a well-flowing order.

In order to reduce the number of free parameters in the model, we grouped species together, and constrained the appropriate competition coefficients to be equal for species within the same group.

A grouping, known as a partition in mathematics, divides the species into groups such that each species belongs to one and only one group.
Given $n$ species, there may be any integer number of groups from $1$ to $n$.
With $n$ groups, the grouping places each species into its own group, such that all species interact in a unique fashion.
With $1$ group, every species is placed in the same group, such that all are equal in their interactions, reminiscent of the neutral model.

We did not assume that species should be grouped the same in their competitive effects as in their competitive responses.
As such, each model was defined by two groupings: a grouping of species constraining equivalence in their competitive effects, and a second, independent grouping of species constraining equivalence in their competitive responses.

The set of competition coefficients in the Lotka--Volterra model can be arranged in an interaction matrix \parencite{novak2016characterizing}, where a row contains all the competitive effects on a single species and a column contains all the competitive effects of a single species.
As such, if two species are in the same group regarding their competitive responses, then two rows of the interaction matrix are constrained to be equal.
Similarly, if two species are in the same group regarding their competitive effects, then two columns of the interaction matrix are constrained to be equal.
This allows us to refer simply to the ``row grouping'' and the ``column grouping''.
Formally, if species $a$ and $b$ were in the same column group, $\alpha_{ia}\equiv\alpha_{ib} \, \forall i$, and if $a$ and $b$ were in the same row group, $\alpha_{aj}\equiv\alpha_{bj} \, \forall j$.
As such, the total number of free competition coefficients in the model was equal to the number of row groups multiplied by the number of column groups.
The species' intrinsic growth rates, $r_i$, were not affected by grouping; each species always had its own growth rate fitted independently of other species.

The total number of groupings for $n$ species is given by the Bell numbers, denoted $B_n$ \parencite{bell1934exponential}.
With six species and independently selected row and column groupings, the total number of possible models is ${\left(B_6\right)}^2 = \num{\Sexpr{numbers::bell(6)^2}}$.

>>>>

@@

The experiment used six species of \emph{Drosophila} from an Australian rainforest community: \emph{D. birchii}, \emph{D. pallidifrons}, \emph{D. pandora}, \emph{D. pseudoananassae}, \emph{D. simulans}, and \emph{D. sulfurigaster}.
The experiment took place in vials with a limited amount of medium for egg-laying and larval development; competition was for space on this medium.
A fixed number of adult pairs of one or two species were added to each vial and allowed to mate and lay eggs.
The number of adult pairs of each species emerging from these eggs was counted.

Experiments occurred for every single-species and two-species combination of the six species, at a variety of densities.
The single-species densities were 3, 9, or 12 pairs per vial, and each of these was replicated six times per species.
The two-species densities were (3,3), (3,6), (3,9), (3,12), (6,3), (9,3), or (12,3) pairs per vial (where the paired values give the densities of each of the two species), and each of these was replicated three times per pair of species.
Vials with a single founding pair were excluded from this study and the original authors' due to an Allee effect that the Lotka--Volterra model cannot accommodate.
Further details of the experimental methodology can be found in \textcite{terry2021natural}.

@@

The data provided the change in population size from one generation to the next, so we used a discrete-time version of the Lotka--Volterra competition model given by:
\begin{equation}
	N_{i,t+1} - N_{i,t} = N_{i,t} \, r_i \left(1 - \sum_j \alpha_{ij} N_{j,t}\right) \,,
	\label{LVmodel}
\end{equation}
where $N_{i,t}$ is the population of species $i$ at time $t$ (the size of the parent generation), $N_{i,t+1}$ is the population at time $t+1$ (the size of the offspring generation), $r_i$ is the intrinsic growth rate of species $i$, and $\alpha_{ij}$ is the per capita effect of species $j$ on species $i$.
The competition coefficients were not constrained to be positive, allowing for the possibility of mutualistic effects between some species.
With $n$ species, this model has $n+n^2$ parameters: $n$ growth rates and $n^2$ competition coefficients.

@@

In order to reduce the number of free parameters in the model, we grouped species together, and constrained the appropriate competition coefficients to be equal for species within the same group.
In our grouping, we avoided the common assumption that species which are similar in their responses to competition would automatically be similar in their competition effects.
As such, we had two groupings; one grouping of the competitive responses, and one of the competitive effects.

The set of competition coefficients in the Lotka--Volterra model can be arranged in an interaction matrix \parencite{novak2016characterizing}, where a row contains all the competitive effects on a single species and a column contains all the competitive effects of a single species.
As such, if two species are assumed to have the same response to competition, then two rows of the interaction matrix are constrained to be equal.
Similarly, if two species are assumed to have the same competitive effect, then two columns of the interaction matrix are constrained to be equal.
This allows us to refer simply to the ``row grouping'' and the ``column grouping''.
Formally, if species $a$ and $b$ were in the same row group, $\alpha_{aj}\equiv\alpha_{bj} \, \forall j$, and if $a$ and $b$ were in the same column group, $\alpha_{ia}\equiv\alpha_{ib} \, \forall i$.
As such, the total number of free competition coefficients in the model was equal to the number of row groups multiplied by the number of column groups.
The species' intrinsic growth rates, $r_i$, were not affected by grouping; each species always had its own growth rate fitted independently of other species.

Groupings are known formally as ``partitions'' in mathematics.
A grouping divides the species into groups such that each species belongs to one and only one group.
Given $n$ species, there may be any integer number of groups from $1$ to $n$.
With $n$ groups, the grouping places each species into its own group, such that all species interact in a unique fashion.
With $1$ group, every species is placed in the same group, such that all are equal in their interactions, reminiscent of the neutral model.
The total number of groupings for $n$ species is given by the Bell numbers, denoted $B_n$ \parencite{bell1934exponential}.
With six species and independently selected row and column groupings, the total number of possible models is ${\left(B_6\right)}^2 = \num{\Sexpr{numbers::bell(6)^2}}$.
